---
title: "Programming a Little App"
author: "Daniel Kaplan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overall organization of a Little App

Little apps are all built on a common template. This enforces visual consistency among the little apps and simplifies programming, since much of a little app works "straight out of the box."



In order to create a new app,

1. Create a directory with a name that will look good as the last part of a URL. That name is how users will access your app.
2. In that directory, place a copy of the `app.R` file from the little apps template directory. You can bring a copy of that file into the RStudio editor by invoking `LA_template_for_app()`:

    ```{r eval = FALSE}
LA_template_for_app()
```

   You will have to save the file as `app.R` in the directory you created in (1).
3. With the `app.R` file open in the editor, press the RStudio "Run App" button and verify that the app runs. At that point you're ready to start customizing the app.

The customization consists of writing your own shiny renderers, reactives, and observers, along with whatever non-reactive code (such as plotting statements and other calculations) needed to support your app.


## Built-in controls

Data sets and controls for choosing response and explanatory variables, and sample size are built in.

Mostly, your reactives will interact with the data via predefined reactives built into the package and template. These include:

* `get_sample()` will return the current sample of data. Only the selected variables will be included.
* `get_sample_size()` gives the current choice for the sample size $n$ as an integer.
* `get_overall_sample_size()`, like `get_sample_size()` but takes into account if the sample is stratified. A stratified sample will have $k \times n$ rows, with $k$ being the number of groups defined by a categorical explanatory variable. (Stratification only applies if the explanatory variable is not  numeric. Stratification is done with replacement, so that low-population groups can  always have $n$ cases in the stratified data.)

The input widgets for the user's choices have standard names, which you should use:

* `input$frame`: the name  of the data frame
* `input$var_y`: name of the response variable, always shown on the y-axis of the graph. (character string)
* `input$var_x`: name of the explanatory variable to be used for the x-axis of the graph. (character string)
* `input$covar`: name(s) of any covariates selected (character string vector with, potentially multiple elements.) If there is no covariate selected, the value is `.none`.
* `input$sample_size`: the  size of the sample. (character string)
* `input$stratify`: whether to stratify the sample according to the explanatory variable (if it's categorical). The total sample size will be the number of groups defined by the explanatory variable times `input$sample_size()`
* `input$new_sample`: triggered when the user presses the new-sample button.

## Default values for uncommonly used inputs

There are some inputs that do not appear in the little-app template, but for which default values are desired. These have been implemented as reactives, that is, you refer to the input value via a reactive. If you want for the non-default value to be available, instantiate these controls in your UI using the names given below.

* `get_spline_order()`: a reactive. If there is an input manual control named `spline_order`, this will report it's value as a number. Otherwise, returns 1 for a straight-line model.
* `include_interaction()` a reactive. If there is an input control named `include_interaction`, this will return its value
* `jitter_width()` a reactive. The corresponding manual control should be `jitter_width`
* `jitter_height()`, like `jitter_width()`.
* `dot_alpha()` sets the alpha level (in a way that depends on the number of points). Corresponding manual control should be named `dot_alpha`

## Built-in outputs

* `output$main_plot`
* `output$codebook <- renderText({ Your HTML })`
* `output$statistics <- renderText({Your HTML})`
* `output$explain <- renderText({Your HTML})`
* `output$rcode <- renderText({Your HTML})`
* To help you debug
    - `output$debug_text <- renderText({Your text})`
    - `output$debug_plot <- renderPlot({Your plot})`
    - `output$debug_table <- renderTable({Your table})`

## Built-in reactives

The data oriented ones


The user interface consists of two main components:

1. A sidebar that will contain choices you make for controls.
2. A main panel, built by `LA_body()` that consists of a shiny `plotOutput()` and a `tabBox()` with several built-in tabs. You can add additional tabs with the `...` argument to `LA_body()`.


```{r echo = FALSE}
body(LA_body)
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
